---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# missForestFast
`missForestFast` is a project for a case study for improving an existing 
random-forst-based imputation method,
[`missForest`](https://CRAN.R-project.org/package=missForest).
The proposed R package added the ability to keep intermediate results
for further study on random-forest-based iterative imputation method.

## Accleration using parallel computation
[`missForestFast`](https://github.com/shangzhi-hong/missForestFast)
provides user with the accelerated version of missing data imputation powered by
[`ranger`](https://CRAN.R-project.org/package=ranger) and [`randomForestSRC`](https://CRAN.R-project.org/package=randomForestSRC)

## Installation for development version from GitHub
Currently the `missForestFast` package is only available as source files on GitHub,
any interested user can install and test the `missForestFast` package by running
the following code in R:

``` r
if(!"remotes" %in% installed.packages()) install.packages("remotes")
remotes::install_github("shangzhi-hong/missForestFast")
```

Run example:
``` r
library(missForestFast)
data(iris)
set.seed(202003)
iris.mis <- prodNA(iris, noNA = 0.5)
impRanger <- missForestRanger(iris.mis, xtrue = iris, keepAll = TRUE)
impSrc <- missForestSrc(iris.mis, xtrue = iris, keepAll = TRUE)
```

## The auto-stop criterion
The auto-stop criterion used by original missForest algorithm may be misleading.
The original missForest algorithm stops the imputation once the out-of-bag error
increases, however, the increase can be instantaneous. As a fact, the symbolic
auto-stop criterion may just be a random stop from the continuous fluctuations
of the out-of-bag error estimates.

Example:
``` r
library(missForestFast)
data(iris)
set.seed(202003)
iris.mis <- prodNA(iris, noNA = 0.25)
targetIter <- 100
impRanger <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = TRUE, forceIter = TRUE)
# Out-of-bag error (for auto-stop)
print(impRanger[["oobErrAll"]])
# Error from true data
print(impRanger[["errAll"]])
```

```{r PlotIterError, echo=FALSE, message=FALSE, warning=FALSE, out.width='50%', fig.align='center'}
library(ggplot2)
library(dplyr)
library(tidyr)
library(missForestFast)
data(iris)
set.seed(202003)
iris.mis <- prodNA(iris, noNA = 0.25)
targetIter <- 100
impRanger <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = T, forceIter = T)
# Out-of-bag error (for auto-stop)
do.call(rbind, impRanger[["oobErrAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
        geom_line() +
        theme_classic() +
        scale_colour_brewer(palette = "Set2") +
        ggtitle("OOB Error")
# Error from true value
do.call(rbind, impRanger[["errAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
    geom_line() +
    theme_classic() +
    scale_colour_brewer(palette = "Set2") +
    ggtitle("Error from True Value")
```


## The initialization of imputation
In our opinion, the missForest algorithm should be recognized as a special case
of MICE (Multivariate Imputation using Chained Equations), using predicted means
as a replacement of samples from conditional distributions.
Use of variable mean (for continuous variable) and most frequent category (for
categorical variable) may not be the best choice.

Example:
``` r
library(missForestFast)
data(iris)
set.seed(202003)
iris.mis <- prodNA(iris, noNA = 0.25)
targetIter <- 100
impRangerMean <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = TRUE, forceIter = TRUE, randInit = FALSE)
# Out-of-bag error with original initialization
print(impRangerMean[["oobErrAll"]])
# Error from true data with original initialization
print(impRangerMean[["errAll"]])
impRangerRand <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = TRUE, forceIter = TRUE, randInit = TRUE)
# Out-of-bag error with random initialization
print(impRangerRand[["oobErrAll"]])
# Error from true data with random initialization
print(impRangerRand[["errAll"]])
```

```{r Init, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width='50%'}
library(ggplot2)
library(dplyr)
library(tidyr)
library(missForestFast)
data(iris)
set.seed(2020)
iris.mis <- prodNA(iris, noNA = 0.20)
targetIter <- 100
impRangerMean <-
    missForestRanger(
        iris.mis,
        xtrue = iris,
        maxiter = targetIter,
        keepAll = TRUE,
        forceIter = TRUE,
        randInit = FALSE
    )
resOobMean <- do.call(rbind, impRangerMean[["oobErrAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Mean")
resTrueMean <- do.call(rbind, impRangerMean[["errAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Mean")
impRangerRand <-
    missForestRanger(
        iris.mis,
        xtrue = iris,
        maxiter = targetIter,
        keepAll = TRUE,
        forceIter = TRUE,
        randInit = TRUE
    )
resOobRand <- do.call(rbind, impRangerRand[["oobErrAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Random")
resTrueRand <- do.call(rbind, impRangerRand[["errAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Random")
bind_rows(resOobMean, resOobRand) %>%
    mutate(Type = paste0(Initialization, "_", Type)) %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
    geom_line() +
    theme_classic() +
    ggtitle("Initialization - OOB Error")

bind_rows(resTrueMean, resTrueRand) %>%
    mutate(Type = paste0(Initialization, "_", Type)) %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
    geom_line() +
    theme_classic() +
    ggtitle("Initialization - True Error")

```


## Number of trees
Even a small number of trees in the random forest model can lead to valid results,
but the results can have unstable out-of-bag errors.

Example:
``` r
library(missForestFast)
data(iris)
set.seed(2020)
iris.mis <- prodNA(iris, noNA = 0.25)
targetIter <- 100
impRangerMean <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = TRUE, forceIter = TRUE, randInit = FALSE, ntree = 100)
# Out-of-bag error with 100 trees
print(impRangerMean[["oobErrAll"]])
# Error from true data with 100 trees
print(impRangerMean[["errAll"]])
impRangerRand <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = TRUE, forceIter = TRUE, randInit = FALSE, ntree = 10)
# Out-of-bag error with 10 trees
print(impRangerRand[["oobErrAll"]])
# Error from true data with 10 trees
print(impRangerRand[["errAll"]])
```

```{r TreeNum, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width='50%'}
library(ggplot2)
library(dplyr)
library(tidyr)
library(missForestFast)
data(iris)
set.seed(2020)
iris.mis <- prodNA(iris, noNA = 0.25)
targetIter <- 100
impRangerMean <-
    missForestRanger(
        iris.mis,
        xtrue = iris,
        maxiter = targetIter,
        keepAll = TRUE,
        forceIter = TRUE,
        randInit = FALSE,
        ntree = 100
    )
resOobMean <- do.call(rbind, impRangerMean[["oobErrAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Tree100")
resTrueMean <- do.call(rbind, impRangerMean[["errAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Tree100")
impRangerRand <-
    missForestRanger(
        iris.mis,
        xtrue = iris,
        maxiter = targetIter,
        keepAll = TRUE,
        forceIter = TRUE,
        randInit = FALSE,
        ntree = 10
    )
resOobRand <- do.call(rbind, impRangerRand[["oobErrAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Tree10")
resTrueRand <- do.call(rbind, impRangerRand[["errAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Tree10")
bind_rows(resOobMean, resOobRand) %>%
    mutate(Type = paste0(Initialization, "_", Type)) %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
    geom_line() +
    theme_classic() +
    ggtitle("Tree Number - OOB Error")

bind_rows(resTrueMean, resTrueRand) %>%
    mutate(Type = paste0(Initialization, "_", Type)) %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
    geom_line() +
    theme_classic() +
    ggtitle("Tree Number - True Error")
```

## Order of variables
By default, missForest algorithm does imputation by the order of missingness
in the variables.
But the order of variables for imputation can be irrelevant to the final results,
even the out-of-bag error can appear higher, but the error from true data shows
the contrary

Example:
``` r
library(missForestFast)
data(iris)
set.seed(2020)
iris.mis <- prodNA(iris, noNA = 0.25)
targetIter <- 100
impRangerDefault <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = TRUE, forceIter = TRUE, randInit = FALSE, ntree = 100, varOrderSeq = FALSE)
# Out-of-bag error with default order
print(impRangerDefault[["oobErrAll"]])
# Error from true data with default order
print(impRangerDefault[["errAll"]])
impRangerSequential <- missForestRanger(iris.mis, xtrue = iris, maxiter = targetIter, keepAll = TRUE, forceIter = TRUE, randInit = FALSE, ntree = 100, varOrderSeq = TRUE)
# Out-of-bag error with sequential order
print(impRangerSequential[["oobErrAll"]])
# Error from true data with sequential order
print(impRangerSequential[["errAll"]])
```
```{r VarOrder, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width='50%'}
library(ggplot2)
library(dplyr)
library(tidyr)
library(missForestFast)
data(iris)
set.seed(2020)
iris.mis <- prodNA(iris, noNA = 0.25)
targetIter <- 100
impRangerDefault <-
    missForestRanger(
        iris.mis,
        xtrue = iris,
        maxiter = targetIter,
        keepAll = TRUE,
        forceIter = TRUE,
        randInit = FALSE,
        ntree = 100,
        varOrderSeq = FALSE
    )
resOobDefault <- do.call(rbind, impRangerMean[["oobErrAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Default")
resTrueDefault <- do.call(rbind, impRangerMean[["errAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Default")

impRangerSequential <-
    missForestRanger(
        iris.mis,
        xtrue = iris,
        maxiter = targetIter,
        keepAll = TRUE,
        forceIter = TRUE,
        randInit = FALSE,
        ntree = 100,
        varOrderSeq = TRUE
    )
resOobSequential <- do.call(rbind, impRangerRand[["oobErrAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Sequential")
resTrueSequential <- do.call(rbind, impRangerRand[["errAll"]]) %>%
    as.data.frame() %>%
    mutate(Iteration = row_number()) %>%
    pivot_longer(c(NRMSE, PFC), names_to = "Type", values_to = "Value") %>%
    mutate(Initialization = "Sequential")
bind_rows(resOobDefault, resOobSequential) %>%
    mutate(Type = paste0(Initialization, "_", Type)) %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
    geom_line() +
    theme_classic() +
    ggtitle("Variable Order - OOB Error")

bind_rows(resTrueDefault, resTrueSequential) %>%
    mutate(Type = paste0(Initialization, "_", Type)) %>%
    ggplot(aes(x = Iteration, y = Value, color = Type)) +
    geom_line() +
    theme_classic() +
    ggtitle("Variable Order - True Error")
```
